Early environmental conditions do not impact associative learning in two species of skink

### 2024-08-30 ms.qmd
Updates

---
title: "Early environmental conditions do not impact associative learning in two species of skink"
docx: 
bibliography: "./bib/refs.bib"
csl: "./bib/animal-cognition.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

## Abstract
**Behavioural flexibility is expected to be crucial for animals adapting to environmental changes. However, the conditions experienced during early development can profoundly impact behavioural flexibility making it unclear how populations will respond to novel circumstances. Stressful situations faced by the parents can have a direct impact on offspring cognition through the transmission of glucocorticoids - stress-related hormones that affect offspring cognition. At the same time, stressful conditions can influence parental behaviour during nesting and consequently the thermal developmental conditions offspring experience. Here, we investigated the interactive effects of prenatal corticosterone (CORT) and temperature on behavioural flexibility in two lizard species, _Lampropholis delicata_ and _L. guichenoti_. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then assessed behavioural flexibility through a reversal learning task. We hypothesized prenatal CORT and cold temperatures to impair with lizards' performance in the reversal task. Given _L. delicata_'s success as an invassive species, we expected this species to behave more flexibly and be less affected by early environmental conditions. Contrary to our expectations, behavioural flexibility appears to be robust to prenatal temperature and CORT in both species. The lack of difference in reversal learning between _L. delicata_ and _L. guichenoti_ suggests that novel environments are unlikely to influence flexible behavioural learning and that behavioural flexibility itself is unlikely to explain differences in invasion success between these species.**

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design of early environment manipulation (**A**) and learning tasks (**B**). Stages 1-3 indicate the different phases of the habituation process. In the associative and reversal tasks, white lids show the ramps where the food reward was not attainable."

@@

#### Statistical analyses  
We performed the analyses for species separately. Because preliminary analyses showed a significant effect of the colour assigned, we decided to split the data by colour (blue or red). Therefore, we run a total of four different Bayesian multilevel models using the brm function from the brms package (@burkner2017brms) in Quarto (@Allaire_Quarto_2022). We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations.

@@

Originally, we started with 96 lizards, 48 per species and 12 per treatment per species. However, due to natural mortality (n = 11), no completion of the training stage (n = 1), or no motivation during the learning tasks (n = 3), we ended up with a total of 81 lizards. Final sample sizes per treatment and species are listed in @fig_deli and @fig-guich (figures for both species with the raw data are included in the Supplementary Material). Mean slopes per treatment for both species are provided in Table 1 in Supplementary Material.  
```{r, models}
#| label: models
# Fitting the model and extraction of posteriors for both species using fit_m function (see func.r in R folder). The result everytime the function is used is a df with the posteriors of the model. The functions saves the model automatically in output/models; and when the parameter refit = FALSE then the posteriors are extracted from the model previously written instead of fitting the model again each time.
source(here("R", "func.R"))
# Model formula: FC_associative ~ Trial*cort*temp + (1 + Trial|lizard_id) + (1| clutch)
## A.1) L. delicata
deli_red <- fit_m("deli", "red", refit = FALSE)
write.csv(deli_red, file= "./output/Checking/deli_red.csv")
deli_blue <- fit_m("deli", "blue", refit = FALSE)
write.csv(deli_blue, file= "./output/Checking/deli_blue.csv")
## A.2) L. guichenoti
guich_red <- fit_m("guich", "red", refit = FALSE)
write.csv(guich_red, file= "./output/Checking/guich_red.csv")
guich_blue <- fit_m("guich", "blue", refit = FALSE)
write.csv(guich_blue, file= "./output/Checking/guich_blue.csv")
```

@@

Because there was a possibility that this bias towards blue feeders was a byproduct of the habituation process, we compared the light spectrum and the perceived chromatic differences between the ramps used in the associative task and the ones used during habituation (see Supplementary Material). Since these results show no similarities between the white and blue ramps, we consider the preference towards blue to not be a derivate from our experimental design. Rather, it seems likely that lizards innately prefer the blue feeders. The species confidence hypothesis posits that animals should bias their attention towards colours they are familiar with [@putman2017fear]. For instance, dark and light blue T-shirts were associated with lower flight initiation distances and higher capture rates in Western fence lizards where blue is used in intraspecific communication [@putman2017fear]. However, to the best of our knowledge, noone has reported a colour bias in _L. delicata_ or _L. guichenoti_ before; and blue colouration is not considered to be involved in intraspecific communication in these species [@torr1996patterns, @chapple_biology_2014]. Still, we recommend to select carefully the colours employed in associative tasks when relying on visual stimuli. Despite this, when lizards assigned to blue or red were analysed separately, the between treatments comparissons remain the same in both conditions and species.

@@

>>>>

---
title: "Early environmental conditions do not impact associative learning in two species of skink"
format:
  docx: 
    reference-doc: "./bib/tmpl.docx"
    number_sections: true
    linestretch: 2
    fontsize: 12pt
    mainfont: Times New Roman
    bibliography: "./bib/refs.bib"
    csl: "./bib/animal-cognition.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

## Abstract
Acquiring new information through learning is essential for animals to adjust their behaviour adaptively. However, animals' abilities to learn can be constrained by the conditions experienced during early stages of development, when the brain is especially susceptible to environmental inputs. Temperature is a pervasive mechanism of phenotypic plasticity in ectotherms, exerting direct effects on growth, metabolism, or learning. In vertebrates, abrupt thermal changes can trigger an allostatic response, which is mediated by glucocorticoids - stress-related hormones - that can be transmitted to the offspring, potentially impacting their learning abilities. Thus, glucocorticoids and the thermal environment are predicted to have interactive effects on the development of learning functions in animals. Here, we investigated the combined consequences of corticosterone (CORT) and temperature elevations on learning in two species of lizards, _Lampropholis delicata_ and _L. guichenoti_. We manipulated prenatal CORT levels and incubation temperature in a 2x2 factorial design, and then subjected juveniles to a colour-associative learning task. In the learning task lizards had to associate the attainability of a food item with the colour of the feeder. We perfomed 35 trials per individual and recorded whether lizards made the correct choice or not. In our analyses, we compared the rate of behavioural change over time (learning slopes) between treatments. We anticipated that elevated CORT and low temperatures would impair lizards' performance. However, our results show that associative learning skills in these two species is robust to prenatal temperature and CORT. We also found a significant effect of the colour employed as the correct choice in our tests for _L. delicata_, underscoring the need to carefully select the color used in cognitive tests involving visual stimuli.

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design of the experiment. Panel **A** shows the early environment manipulation. In panel **B**, the measurements of the 3D-printed ramps employed in the habituation and learning tasks. Panel **C** shows the habituation process with the three different stages. And in panel **D**, the associative task. White lids in **D** show the ramps where the food reward was not attainable."

@@

#### Statistical analyses  
We performed the analyses for species separately. Because preliminary analyses showed a significant effect of the colour assigned, we decided to split the data by colour (blue or red). Therefore, we run a total of four different Bayesian multilevel models using the brm function from the brms package [@burkner2017brms] in Quarto [@Allaire_Quarto_2022]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations.

@@

Originally, we started with 96 lizards, 48 per species and 12 per treatment per species. However, due to natural mortality (n = 11), no completion of the training stage (n = 1), or no motivation during the learning tasks (n = 3), we ended up with a total of 81 lizards. Final sample sizes per treatment and species are listed in @fig-deli and @fig-guich (figures for both species with the raw data are included in the Supplementary Material). Mean slopes per treatment for both species are provided in Table 1 in Supplementary Material.  
```{r, models}
#| label: models
# Fitting the model and extraction of posteriors for both species using fit_m function (see func.r in R folder). The result everytime the function is used is a df with the posteriors of the model. The functions saves the model automatically in output/models; and when the parameter refit = FALSE then the posteriors are extracted from the model previously written instead of fitting the model again each time.
source(here("R", "func.R"))
# Model formula: FC_associative ~ Trial*cort*temp + (1 + Trial|lizard_id) + (1| clutch)
## A.1) L. delicata
deli_red <- fit_m("deli", "red", refit = FALSE)
write.csv(deli_red, file= "./output/Checking/deli_red.csv")
deli_blue <- fit_m("deli", "blue", refit = FALSE)
write.csv(deli_blue, file= "./output/Checking/deli_blue.csv")
## A.2) L. guichenoti
guich_red <- fit_m("guich", "red", refit = FALSE)
write.csv(guich_red, file= "./output/Checking/guich_red.csv")
guich_blue <- fit_m("guich", "blue", refit = FALSE)
write.csv(guich_blue, file= "./output/Checking/guich_blue.csv")
```

@@

Because there was a possibility that this bias towards blue feeders was a byproduct of the habituation process, we compared the light spectrum and the perceived chromatic differences between the ramps used in the associative task and the ones used during habituation (see Supplementary Material). Since these results show no similarities between the white and blue ramps, we consider the preference towards blue to not be a derivate from our experimental design. Rather, it seems likely that lizards innately prefer the blue feeders. The species confidence hypothesis posits that animals should bias their attention towards colours they are familiar with [@putman2017fear]. For instance, dark and light blue T-shirts were associated with lower flight initiation distances and higher capture rates in Western fence lizards where blue is used in intraspecific communication [@putman2017fear]. However, to the best of our knowledge, noone has reported a colour bias in _L. delicata_ or _L. guichenoti_ before; and blue colouration is not considered to be involved in intraspecific communication in these species [@torr1996patterns; @chapple_biology_2014]. Still, we recommend to select carefully the colours employed in associative tasks when relying on visual stimuli. Despite this, when lizards assigned to blue or red were analysed separately, the between treatments comparissons remain the same in both conditions and species.

@@

## Data accessibility
All data, data description, and R code are available in public repository [https://github.com/Pablo-Recio/CORT_Temp_learning](https://github.com/Pablo-Recio/CORT_Temp_learning).

## Satements and declarations 
#### Competing Interests
We declare we have no competing interests
#### Conflict of Interest
We declare we have no conflict of interest
#### Ethical Approval
All experimental procedures and laboratory housing complied with Australian law and were approved by the Australian National University Animal Experimentation Ethics Committee (A2022_33).

## Authors’ contributions
P.R.: conceptualization, methodology, data collection, data curation, formal analysis, writing—original draft, writing—review and editing; D.C.L.: data collection, writing—review and editing; O.C.: conceptualization, methodology, writing—review and editing; K.H.W.: conceptualization, writing—review and editing; C.F.: conceptualization, methodology, funding acquisition, writing—review and editing; B.M: data collection, writing—review and editing; A.Y.P.: data collection, writing—review and editing; D.N.: conceptualization, methodology, funding acquisition, project administration, resources, supervision, writing—review and editing.
All authors gave final approval for publication and agreed to be held accountable for the work performed therein.

## Funding
This work was supported by a National Australian University fellowship (P.R.), the Australian Research Council (grant no. DP210101152) to D.N. and C.F., and the ACT Herpetological Association grant to P.R.

## Acknowledgements
We thank the help and assistance of our lab technicians Benjamin Durant and Michelle Stephens for taking care of the lizards. We are also grateful to ACTHA for the grant for the 3D printed feeders, and we also thank ANU MakerSpace, where we designed and built the prototypes of the 3D printed feeders. Finally, we wish to acknowledge the anonymous reviewers for their valuable feedback on the manuscript.
